From Reversible Programs to \\ Univalent Universes and Back

### 2021-07-08 Pi+/paper/ufin.tex
working on ufin

% First - recap the previous section
% Second - say what was wrong
To recap, the denotational semantics of $\PiLang$ interprets types as objects in $\BFin$, 1-combinators as isomorphisms,
and we prove that for every pair of 1-combinators related by a 2-combinator, their interpretations in~$\BFin$ are equal.
The groupoid $\BFin$ is strict, since the collection of isomorphisms is a set, that is, a discrete category. There is no
explicit witness for the equality of two isomorphisms, since we can decide by evaluating two bijections whether they're
equal. To be able to establish completeness for $\PiLang$, we want a witness for this equality, so that we can quote
back to the syntax and produce a 2-combinator witnessing the equality of the corresponding 1-combinators.

@@

% Third - say how are we fixing it
First, we observe that the implicit equalities between the isomorphisms are pointwise equalities of functions, that is,
homotopies -- we \emph{weaken} the groupoid $\BFin$, exposing these homotopies, by using higher invertible cells for
them. We work in Homotopy Type Theory (Univalent Foundations) -- it provides a proof-relevant, constructive metatheory
to get a handle on these equalities, and, it provides a rich internal language for describing weak groupoids, using the
"types are weak $\infty$-groupoids" correspondence.

@@

% Fourth - expain the tool we use for fixing it
Every type in HoTT is a weak $\infty$-groupoid whose points are the terms of the type, and the (iterated) identity type
gives the (higher) morphisms. The groupoid we're interested in has types as points, type equivalences for 1-cells, and
higher homotopies for higher cells. This is the groupoid structure for the universe type $\UU$, since the identity type
on types can be characterised as type equivalences (by \emph{univalence}). But, we only want to carve out a
\emph{subuniverse} of \emph{finite types}, still satisfying univalence, to get the groupoid structure. In this section,
we formally define \emph{univalent subuniverses}, and proceed to construct the particular instance for finite types,
$\UFin$ (\Cref{def:ufin}).

@@

% In HoTT, every type carries a groupoid structure, with terms for points, and the (iterated) identity type for (higher)
% morphisms. We're interested in the groupoid structure of the universe, which has types for points, equivalences for
% identities between them (by univalence), and higher homotopies for higher identitifications. We want a similar
% universe, whose only points are the finite types, that is, a subuniverse, and, equalities between them should
% correspond to equivalences of finite types -- the subuniverse should be univalent. To construct a univalent
% subuniverse starting from a univalent universe, we will introduce the concept of univalent fibrations.

@@

\begin{toappendix}
  We give an example of the groupoid structure on a 3-element set.

@@

  We have $\Fin[3] = \Set{0,1,2} \eqv \unit \sqcup (\unit \sqcup \unit)$ which fixes a particular enumeration of the
  elements. Suppose we have a set $X = (\unit \sqcup \unit) \sqcup \unit$, it has the same cardinality as $\Fin[3]$, so it
  is represented by the same 0-cell. But, $X$ can be made equivalent to $\Fin[3]$ in many different ways since there are
  many bijections between them. One bijection is
  $\Set{\inl(\inl(\ttt)) \mapsto 0, \inl(\inr(\ttt)) \mapsto 1, \inr(\ttt) \mapsto 2}$ which can be written in two
  different ways by composing more primitive operations, $f_{1} = \assocrp$, or
  $f_{2} = \swapp \compc \assoclp \compc \swapp$. Another bijection is
  $\Set{\inl(\inl(\ttt)) \mapsto 1, \inl(\inr(\ttt)) \mapsto 2, \inr(\ttt) \mapsto 0}$ which is given by $f_{3} = \ldots$.
  Since $f_{1}$ and $f_{2}$ produce the same enumeration of the elements of $X$, they are identified by a homotopy $h$
  which is encoded in the 2-cell between them.

  At level 0, all we know is that if $X : \UFin[3]$, then X is merely equal to $\Fin[3]$, that is
  $\Trunc[-1]{X \id \Fin[3]}$, and we don't have access to the bijection. At level 1, if we know that both $X$ and $Y$ are
  \emph{equal} in $\UFin[3]$, then we can extract an equivalence between them, that is, $(X \id Y) \to (X \eqv Y)$.
  $\UFin[3]$ being a univalent subuniverse asserts that there are as many elements (upto higher homotopy) in $X \id Y$ as
  there are $X \eqv Y$.

  \subsection{To Put Somewhere or Delete}

  This approach is sufficient to prove the semantics forms a 1-category but ignores the rich structure at the next
  level~\cite{carette2016}.
  As explained in the previous section, a $\Pi$-type $A$ has $\sizet{A}$-elements and for all combinators $c : A \iso B$
  we have that $\sizet{A} = \sizet{B}$. Hence, the denotation $\denot{A}$ of a type $A$ with $n$-elements can be the finite
  set $\Fin[n] = \{ 0, 1, \cdots, n-1\}$; the denotation of a value $v : A$ such that $\sizet{A}=n$ will be an index in
  the range $[0,n-1]$, and the denotation $\denot{c}$ of a combinator $c : A \iso B$ such that
  $\sizet{A} = \sizet{B} = n$ will be a function from $\Fin[n]$ to $\Fin[n]$ that permutes the elements. Thus, all types
  with 3 elements will denote $\Fin[3]$ and combinators between them will denote permutations on $\Fin[3]$, e.g.:
  \[\begin{array}{rcl}
      \denot{\onet + (\onet + \onet)}                                         & = & \{ 0,1,2 \} \\
      \denot{(\onet + \onet) + \onet}                                         & = & \{ 0,1,2 \} \\
      \\
      \denot{\assoclp : \onet + (\onet + \onet) \iso (\onet + \onet) + \onet} & = & (0~1~2)     \\
      \denot{\swapp : \onet + (\onet + \onet) \iso (\onet + \onet) + \onet}   & = & (2~0~1)
    \end{array}\]
  where we have used the one-line notation for permutations with $(a~b~c)$ representing the
  permutation that maps 0 to $a$, 1 to $b$, and 2 to $c$. To make the denotation of values precise, we compute a canonical
  enumeration of the elements of each type:
  \[\begin{array}{rcl}
      \mathit{enum}(\zerot)     & = & [~]                                                                                                          \\
      \mathit{enum}(\onet)      & = & [ ~\Acon{tt}~ ]                                                                                              \\
      \mathit{enum}(A + B)      & = & \mathit{map}~\Acon{inj₁}~\mathit{enum}(A) ~\textsf{+\!+}~ \mathit{map}~\Acon{inj₂}~\mathit{enum}(B)          \\
      \mathit{enum}(A \times B) & = & \mathit{concat}~(\mathit{map}~(\lambda v.\mathit{map}~(\lambda w. (v,w))~\mathit{enum}(B))~\mathit{enum}(A))
    \end{array}\]
  \noindent The specification uses a Haskell-like notation for sequences with $\mathit{map}$ as the operation that applies
  a function to each element of a sequence, \textsf{+\!+} as the binary append operation, and $\mathit{concat}$ the
  operation that appends all the subsequences in a sequence.

  Using the definition, we have:
  \[\begin{array}{rcl}
      \mathit{enum}(\onet + (\onet + \onet)) & = & [ \inlv{\Acon{tt}},~\inrv{(\inlv{\Acon{tt}})},~\inrv{(\inrv{\Acon{tt}})} ] \\
      \mathit{enum}((\onet + \onet) + \onet) & = & [ \inlv{(\inlv{\Acon{tt}})},~\inlv{(\inrv{\Acon{tt}})},~\inrv{\Acon{tt}} ]
    \end{array}\]
  Thus, as shown in the diagrams below, $\assoclp~(\inlv{\Acon{tt}})$ applies the permutation $(0~1~2)$ to the index of
  $\inlv{\Acon{tt}}$ which is 0 and produces index 0 in the $(\onet + \onet) + \onet$ type corresponding to value
  $\inlv{(\inlv{\Acon{tt}})}$. Similarly, $\swapp~(\inlv{\Acon{tt}})$ applies the permutation $(2~0~1)$ to the index of
  $\inlv{\Acon{tt}}$ which is 0 and produces index 2 in the $(\onet + \onet) + \onet)$ type corresponding to value
  $\inrv{\Acon{tt}}$.

  \begin{center}
    \input{assoc-perm.tikz}
    \qquad
    \input{swap-perm.tikz}
  \end{center}

  We choose a canonical set of size $n$, called $\mathsf{Fin}~n$, whose elements are natural numbers less than $n$. To
  compute the denotation of a type $A$, we first calculate its size $n = \sizet{A}$. We then construct the canonical set
  $\mathsf{Fin}~n$ and provide the (trivial) evidence that this set is identical to $(\mathsf{Fin}~n)$:

  \[\begin{array}{rcll}
      \sem{A} & = & (\mathsf{Fin}~n, [ n , \mathsf{refl} ]) & \mbox{where}~\sizet{A} = n
    \end{array}\]

  \noindent The denotation $\sem{c}$ of a combinator $c : A \isoone B$ is a path between $\sem{A}$ and $\sem{B}$. If the
  size of $A$ is $m$ and the size of $B$ is $n$, the desired path is between $(\mathsf{Fin}~m, [ m , \mathsf{refl} ])$ and
  $(\mathsf{Fin}~n, [ n , \mathsf{refl} ])$. This path is directly constructed using $\mathit{ap}$ and the fact that $m=n$
  since combinators are always between types of the same size.

  \noindent Finally, given two combinators $p , q : A \isoone B$ and a 2-combinator $\alpha : p \isotwo q$, the denotation
  $\sem{\alpha}$ of $\alpha$ is a path between $\sem{p}$ and $\sem{q}$.

  \note{We use the rig structure of $\UFin$ in~\Cref{subsec:rig} to interpret $\PiLang$.}

  We need a formal definition of normal form (canonical form)

  Recalling that the $\lambda$-calculus arises as the internal language of Cartesian Closed Categories
  (Elliott~\cite{Elliott-2017} gives a particularly readable account of this), we can think of $\Pi$ in similar terms, but
  for symmetric Rig Groupoids instead. For example, we can ask what does the equivalence above represent? It is actually a
  ``linear'' representation of a 2-categorial commutative diagram! In fact, it is a painfully verbose version thereof, as
  it includes many \emph{refocusing} steps because our language does not build associativity into its syntax. Categorical
  diagrams usually do.  Thus if we rewrite the example in diagrammatic form, eliding all uses of associativity, but
  keeping explicit uses of identity transformations, we get that \AgdaFunction{swap{-}fl2⇔swap{-}fl1} represents

  \vspace*{3mm}
  \begin{tikzcd}[column sep=normal, row sep=normal]
    && (a+c)+b \arrow [r, "\swapp \oplus\idd", ""{name=U, below}] & (c+a)+b \arrow [dr, "\assocrp"] && \\
    & a+(c+b) \arrow [ur, "\assoclp"] & & & c+(a+b) \arrow [dr, "\idd\oplus\swapp"] &  \\
    a+(b+c) \arrow [ur, "\idd\oplus\swapp"] \arrow [r, "\assoclp"]
    \arrow [dr, "\assoclp"]
    \arrow [ddr, swap, "\assoclp"]
    & (a+b)+c \arrow [r, "\swapp"] &
    c+(a+b) \arrow [r, swap, "\assoclp", ""{name=D, above}]
    & |[alias=Z]| (c+a)+b \arrow [r, "\assocrp"] &c+(a+b) \arrow [r, "\idd\oplus\swapp"] & c+(b+a) \\
    & (a+b)+c \arrow [dr, "\swapp"] &&&& \\
    & (a+b)+c \arrow [dr, swap, "\swapp"] & c+(a+b) \arrow [rr, swap, "\idd", ""{name=DD, above}]
    \arrow [d, Rightarrow, "\idf\, \mathit{idl}\odot{l}"] &&
    c+(a+b) \arrow [ruu, "\idd\oplus\swapp"] & \\
    && c+(a+b) \arrow [rrruuu, bend right = 40, swap, "\idd\oplus\swapp"] && \\
    \arrow[Rightarrow, from=U, to=D, "\mathit{hexagon}\oplus{r}\, \boxdot\, \idf"]
    \arrow[Rightarrow, from=Z, to=DD, swap, "\idf\boxdot\mathit{linv}\odot{l}\,\boxdot\,\idf"]
  \end{tikzcd}

@@

We use the type theory of the HoTT book~\cite{univalentfoundationsprogramHomotopyTypeTheory2013}, that is, we use
intensional Martin-L\"{o}f Type Theory, with a (univalent) universe $\UU$, and a few Higher Inductive Types (HITs) for
propositional and set truncation, and set-quotients. All arguments will hold in a Cubical Type Theory as well, such
as~\cite*{cohenCubicalTypeTheory2018,angiuliComputationalSemanticsCartesianCubical2019,vezzosiCubicalAgdaDependently2019}.
In the appendix, we review some basic concepts and notation from the HoTT book that we use. For more details, we refer
the reader to the book~\cite{univalentfoundationsprogramHomotopyTypeTheory2013}.

@@

\begin{toappendix}
  \subsubsection{Identity Types}

@@

  A homotopy between functions $f \htpy g$ is given by pointwise equality between them $\dfun{x:A}{f(x) \id_{B} g(x)}$.
  The identity type for functions is equivalent to homotopies between them ${(f \id_{A \to B} g)} \eqv {(f \htpy g)}$, by
  function extensionality. An equivalence between types $A \eqv B$ is given by a pair of functions between them which
  compose to the identity, $f \comp g \htpy \idfunc_{B}$ and $g \comp f \htpy \idfunc_{A}$, and this is equivalent to the
  identity type for the universe, $(A \id_{\UU} B) \eqv (A \eqv B)$, by univalence.

@@

  The type $\dsum*{x:A}{P(x)}$ is the collection of all the fibers and is called the total space of $P$. The first
  projection ${\pi_1 : \dsum*{x:A}{P(x)} \to A}$ from the total space to the base space $A$ has the structure of a
  fibration, that is, there is a lifting operation (\Cref{fig:lift}) which lifts paths in the base space to paths in the
  total space. Given a path $p : x \id_{A} y$ in the base space, and $u : P(x)$ a point in the fiber over $x$, we have

@@

Functions between types are functors between groupoids, and type families, or functions to the universe, are indexed
families of groupoids. A type family $P : A \to \UU$ comes equipped with a functor $\pi_1 : \dsum*{x:A}{P(x)} \to \UU$
which has a lifting operation giving it the structure of a fibration (see the appendix for details). The
$\term{transport}$ operation lifts paths in the base space to maps between the fibers. Using the groupoid structure of
$A$, we can show that $\transport{P}{p}{\blank}$ and $\transport{P}{\inv{p}}{\blank}$ form an equivalence -- we name
this map $\tptEqv{P}$ which lifts paths to equivalences of fibers.

\[
  \tptEqv{P} : \dfun{x,y:A}{x \id_{A} y \to P(x) \eqv P(y)}
\]

The type families (or fibrations) we're interested in are the ones where paths in the base space completely determine
the equivalences in the fibers -- these are called univalent
fibrations~\cite*{kapulkinUnivalenceSimplicialSets2018,kapulkinSimplicialModelUnivalent2021,christensenCharacterizationUnivalentFibrations2015}.

@@

  \subsubsection{Homotopy Types}

@@

  \subsubsection{Higher Inductive Types}

@@

  \begin{definition}[Set Quotient]
    \label{def:set-quot}
    Given a type $A$ which is an $\hSet$, and a relation $R : A \to A \to \hProp$, the set-quotient $\quot{A}{R}$ is the
    higher inductive type generated by
    \begin{itemize}
      \item an inclusion function $q : A \to \quot{A}{R}$,
      \item for each $x, y : A$ such that $R(x,y)$, a path $q(x) \id_{\quot{A}{R}} q(y)$,
      \item a set truncation, for each $x, y : \quot{A}{R}$ and $r, s : x \id_{\quot{A}{R}} y$, we have $r \id s$,
    \end{itemize}
    with \review{an appropriate induction principle.}
  \end{definition}

@@

\begin{proposition}[Univalent Subuniverse]
  A universe predicate is a type family $P : \UU \to \UU$ whose fibers are propositions, that is, $P(X)$ is a
  proposition for every $X$. The fibration ${\fst : \dsum*{X:\UU}{P(X)} \to \UU}$ is univalent. This generates a
  univalent subuniverse
  \[
    \UU_{P} \defeq (\dsum{X:\UU}{P(X)}, \fst)
  \]
\end{proposition}

\begin{appendixproof}
  Suppose $(U, \El) \defeq (\dsum*{X:\UU}{P(X)}, \fst)$ is a subuniverse generated by a subtype $P : \UU \to \UU$. For
  any $X, Y : \UU$ such that $\phi : P(X)$ and $\psi : P(Y)$, we want to show that
  $\tptEqv{\fst} : (X,\phi) \id (Y,\psi) \to X \eqv Y$ is an equivalence. We construct
  $X \eqv Y \to (X,\phi) \id (Y,\psi)$ by $\ua$ and using the fact that $P(\blank)$ is a proposition. That it is an
  inverse follows by calculation using the appropriate computation rules.
\end{appendixproof}

The types we're interested in are the finite types. In constructive mathematics, the notion of finiteness is
subtle~\cite{spiwackConstructivelyFinite2010}. We use the notion of Bishop-finiteness -- a type is finite if it is
merely equivalent to a finite set (\Cref{def:finite-set}), and we define the predicate $\isFin$ (\Cref{def:isfin}) to do
so.

>>>>

% First - recap the previous section
% Second - say what was wrong

Using categorical language, the setting for the semantics in the previous section is the category of finite sets and
functions $\SetFin$. However, as $\PiLang$ only refers to bijective functions, a more precise setting is the groupoid
$\BFin = \term{core}(\SetFin)$ of finite sets and bijections. $\SetFin$ has finite coproducts $(\emptyt, \sqcup)$ and
finite products $(\unit, \times)$ and in $\BFin$ these restrict to additive and multiplicative symmetric monoidal
structures, respectively, making $\BFin$ a symmetric rig groupoid.

The semantics interprets $\PiLang$-types as objects in $\BFin$, 1-combinators as isomorphisms, and for every pair of 1-combinators
related by a 2-combinator, their interpretations in~$\BFin$ are equal.  The groupoid $\BFin$ is strict, since the
collection of isomorphisms is a set, that is, a discrete category. There is no explicit witness for the equality of two
isomorphisms, since we can decide by evaluating two bijections whether they are equal. To be able to establish
completeness for $\PiLang$, we want a witness for this equality, so that we can quote back to the syntax and produce a
2-combinator witnessing the equality of the corresponding 1-combinators.

@@

% Third - say how are we fixing it

We observe that the implicit equalities between the isomorphisms are pointwise equalities of functions, that is,
homotopies. We therefore \emph{weaken} the groupoid $\BFin$, exposing these homotopies, by using higher invertible cells
and work in HoTT (Univalent Foundations) as it provides a proof-relevant, constructive metatheory to get a handle
on these equalities and provides a rich internal language for describing weak groupoids, using the ``types are weak
$\infty$-groupoids'' correspondence.

@@

% Fourth - explain the tool we use for fixing it
Every type in HoTT is a weak $\infty$-groupoid whose points are the terms of the type, and the (iterated) identity type
gives the (higher) morphisms. The groupoid we are interested in has types as points, type equivalences for 1-cells, and
higher homotopies for higher cells. (See App.~\ref{app:grpdexample} for an example on a 3-element set.) This is the
groupoid structure for the universe type $\UU$, since the identity type on types can be characterised as type
equivalences (by \emph{univalence}). But, we only want to carve out a \emph{subuniverse} of \emph{finite types}, still
satisfying univalence, to get the groupoid structure. In this section, we formally define \emph{univalent subuniverses},
and proceed to construct the particular instance for finite types, $\UFin$ (\Cref{def:ufin}).

@@

% In HoTT, every type carries a groupoid structure, with terms for points, and the (iterated) identity type for (higher)
% morphisms. We're interested in the groupoid structure of the universe, which has types for points, equivalences for
% identities between them (by univalence), and higher homotopies for higher identifications. We want a similar
% universe, whose only points are the finite types, that is, a subuniverse, and, equalities between them should
% correspond to equivalences of finite types -- the subuniverse should be univalent. To construct a univalent
% subuniverse starting from a univalent universe, we will introduce the concept of univalent fibrations.

@@

\begin{toappendix}
  \label{app:grpdexample}
  We give an example of the groupoid structure on a 3-element set.

@@

  \noindent We have $\Fin[3] = \Set{0,1,2} \eqv \unit \sqcup (\unit \sqcup \unit)$ which fixes a particular enumeration of the
  elements. Suppose we have a set $X = (\unit \sqcup \unit) \sqcup \unit$, it has the same cardinality as $\Fin[3]$, so it
  is represented by the same 0-cell. But, $X$ can be made equivalent to $\Fin[3]$ in many different ways since there are
  many bijections between them. One bijection is
  $\Set{\inl(\inl(\ttt)) \mapsto 0, \inl(\inr(\ttt)) \mapsto 1, \inr(\ttt) \mapsto 2}$ which can be written in two
  different ways by composing more primitive operations, $f_{1} = \assocrp$, or
  $f_{2} = \swapp \compc \assoclp \compc \swapp$. Another bijection is
  $\Set{\inl(\inl(\ttt)) \mapsto 1, \inl(\inr(\ttt)) \mapsto 2, \inr(\ttt) \mapsto 0}$ which is given by $f_{3} = \ldots$.
  Since $f_{1}$ and $f_{2}$ produce the same enumeration of the elements of $X$, they are identified by a homotopy $h$
  which is encoded in the 2-cell between them.

  At level 0, all we know is that if $X : \UFin[3]$, then X is merely equal to $\Fin[3]$, that is
  $\Trunc[-1]{X \id \Fin[3]}$, and we don't have access to the bijection. At level 1, if we know that both $X$ and $Y$ are
  \emph{equal} in $\UFin[3]$, then we can extract an equivalence between them, that is, $(X \id Y) \to (X \eqv Y)$.
  $\UFin[3]$ being a univalent subuniverse asserts that there are as many elements (upto higher homotopy) in $X \id Y$ as
  there are $X \eqv Y$.
\end{toappendix}

% \begin{toappendix}

%   \subsection{To Put Somewhere or Delete}

%   This approach is sufficient to prove the semantics forms a 1-category but ignores the rich structure at the next
%   level~\cite{carette2016}.
%   As explained in the previous section, a $\Pi$-type $A$ has $\sizet{A}$-elements and for all combinators $c : A \iso B$
%   we have that $\sizet{A} = \sizet{B}$. Hence, the denotation $\denot{A}$ of a type $A$ with $n$-elements can be the finite
%   set $\Fin[n] = \{ 0, 1, \cdots, n-1\}$; the denotation of a value $v : A$ such that $\sizet{A}=n$ will be an index in
%   the range $[0,n-1]$, and the denotation $\denot{c}$ of a combinator $c : A \iso B$ such that
%   $\sizet{A} = \sizet{B} = n$ will be a function from $\Fin[n]$ to $\Fin[n]$ that permutes the elements. Thus, all types
%   with 3 elements will denote $\Fin[3]$ and combinators between them will denote permutations on $\Fin[3]$, e.g.:
%   \[\begin{array}{rcl}
%       \denot{\onet + (\onet + \onet)}                                         & = & \{ 0,1,2 \} \\
%       \denot{(\onet + \onet) + \onet}                                         & = & \{ 0,1,2 \} \\
%       \\
%       \denot{\assoclp : \onet + (\onet + \onet) \iso (\onet + \onet) + \onet} & = & (0~1~2)     \\
%       \denot{\swapp : \onet + (\onet + \onet) \iso (\onet + \onet) + \onet}   & = & (2~0~1)
%     \end{array}\]
%   where we have used the one-line notation for permutations with $(a~b~c)$ representing the
%   permutation that maps 0 to $a$, 1 to $b$, and 2 to $c$. To make the denotation of values precise, we compute a canonical
%   enumeration of the elements of each type:
%   \[\begin{array}{rcl}
%       \mathit{enum}(\zerot)     & = & [~]                                                                                                          \\
%       \mathit{enum}(\onet)      & = & [ ~\Acon{tt}~ ]                                                                                              \\
%       \mathit{enum}(A + B)      & = & \mathit{map}~\Acon{inj₁}~\mathit{enum}(A) ~\textsf{+\!+}~ \mathit{map}~\Acon{inj₂}~\mathit{enum}(B)          \\
%       \mathit{enum}(A \times B) & = & \mathit{concat}~(\mathit{map}~(\lambda v.\mathit{map}~(\lambda w. (v,w))~\mathit{enum}(B))~\mathit{enum}(A))
%     \end{array}\]
%   \noindent The specification uses a Haskell-like notation for sequences with $\mathit{map}$ as the operation that applies
%   a function to each element of a sequence, \textsf{+\!+} as the binary append operation, and $\mathit{concat}$ the
%   operation that appends all the subsequences in a sequence.

%   Using the definition, we have:
%   \[\begin{array}{rcl}
%       \mathit{enum}(\onet + (\onet + \onet)) & = & [ \inlv{\Acon{tt}},~\inrv{(\inlv{\Acon{tt}})},~\inrv{(\inrv{\Acon{tt}})} ] \\
%       \mathit{enum}((\onet + \onet) + \onet) & = & [ \inlv{(\inlv{\Acon{tt}})},~\inlv{(\inrv{\Acon{tt}})},~\inrv{\Acon{tt}} ]
%     \end{array}\]
%   Thus, as shown in the diagrams below, $\assoclp~(\inlv{\Acon{tt}})$ applies the permutation $(0~1~2)$ to the index of
%   $\inlv{\Acon{tt}}$ which is 0 and produces index 0 in the $(\onet + \onet) + \onet$ type corresponding to value
%   $\inlv{(\inlv{\Acon{tt}})}$. Similarly, $\swapp~(\inlv{\Acon{tt}})$ applies the permutation $(2~0~1)$ to the index of
%   $\inlv{\Acon{tt}}$ which is 0 and produces index 2 in the $(\onet + \onet) + \onet)$ type corresponding to value
%   $\inrv{\Acon{tt}}$.

%   \begin{center}
%     \input{assoc-perm.tikz}
%     \qquad
%     \input{swap-perm.tikz}
%   \end{center}

%   We choose a canonical set of size $n$, called $\mathsf{Fin}~n$, whose elements are natural numbers less than $n$. To
%   compute the denotation of a type $A$, we first calculate its size $n = \sizet{A}$. We then construct the canonical set
%   $\mathsf{Fin}~n$ and provide the (trivial) evidence that this set is identical to $(\mathsf{Fin}~n)$:

%   \[\begin{array}{rcll}
%       \sem{A} & = & (\mathsf{Fin}~n, [ n , \mathsf{refl} ]) & \mbox{where}~\sizet{A} = n
%     \end{array}\]

%   \noindent The denotation $\sem{c}$ of a combinator $c : A \isoone B$ is a path between $\sem{A}$ and $\sem{B}$. If the
%   size of $A$ is $m$ and the size of $B$ is $n$, the desired path is between $(\mathsf{Fin}~m, [ m , \mathsf{refl} ])$ and
%   $(\mathsf{Fin}~n, [ n , \mathsf{refl} ])$. This path is directly constructed using $\mathit{ap}$ and the fact that $m=n$
%   since combinators are always between types of the same size.

%   \noindent Finally, given two combinators $p , q : A \isoone B$ and a 2-combinator $\alpha : p \isotwo q$, the denotation
%   $\sem{\alpha}$ of $\alpha$ is a path between $\sem{p}$ and $\sem{q}$.

%   \note{We use the rig structure of $\UFin$ in~\Cref{subsec:rig} to interpret $\PiLang$.}

%   We need a formal definition of normal form (canonical form)

%   Recalling that the $\lambda$-calculus arises as the internal language of Cartesian Closed Categories
%   (Elliott~\cite{Elliott-2017} gives a particularly readable account of this), we can think of $\Pi$ in similar terms, but
%   for symmetric Rig Groupoids instead. For example, we can ask what does the equivalence above represent? It is actually a
%   ``linear'' representation of a 2-categorial commutative diagram! In fact, it is a painfully verbose version thereof, as
%   it includes many \emph{refocusing} steps because our language does not build associativity into its syntax. Categorical
%   diagrams usually do.  Thus if we rewrite the example in diagrammatic form, eliding all uses of associativity, but
%   keeping explicit uses of identity transformations, we get that \AgdaFunction{swap{-}fl2⇔swap{-}fl1} represents

%   \vspace*{3mm}
%   \begin{tikzcd}[column sep=normal, row sep=normal]
%     && (a+c)+b \arrow [r, "\swapp \oplus\idd", ""{name=U, below}] & (c+a)+b \arrow [dr, "\assocrp"] && \\
%     & a+(c+b) \arrow [ur, "\assoclp"] & & & c+(a+b) \arrow [dr, "\idd\oplus\swapp"] &  \\
%     a+(b+c) \arrow [ur, "\idd\oplus\swapp"] \arrow [r, "\assoclp"]
%     \arrow [dr, "\assoclp"]
%     \arrow [ddr, swap, "\assoclp"]
%     & (a+b)+c \arrow [r, "\swapp"] &
%     c+(a+b) \arrow [r, swap, "\assoclp", ""{name=D, above}]
%     & |[alias=Z]| (c+a)+b \arrow [r, "\assocrp"] &c+(a+b) \arrow [r, "\idd\oplus\swapp"] & c+(b+a) \\
%     & (a+b)+c \arrow [dr, "\swapp"] &&&& \\
%     & (a+b)+c \arrow [dr, swap, "\swapp"] & c+(a+b) \arrow [rr, swap, "\idd", ""{name=DD, above}]
%     \arrow [d, Rightarrow, "\idf\, \mathit{idl}\odot{l}"] &&
%     c+(a+b) \arrow [ruu, "\idd\oplus\swapp"] & \\
%     && c+(a+b) \arrow [rrruuu, bend right = 40, swap, "\idd\oplus\swapp"] && \\
%     \arrow[Rightarrow, from=U, to=D, "\mathit{hexagon}\oplus{r}\, \boxdot\, \idf"]
%     \arrow[Rightarrow, from=Z, to=DD, swap, "\idf\boxdot\mathit{linv}\odot{l}\,\boxdot\,\idf"]
%   \end{tikzcd}

% \end{toappendix}

@@

We use the type theory of the HoTT book~\cite{univalentfoundationsprogramHomotopyTypeTheory2013}, that is, we use
intensional Martin-L\"{o}f Type Theory, with a (univalent) universe $\UU$, and a few Higher Inductive Types (HITs) for
propositional and set truncation, and set-quotients. All arguments will hold in a Cubical Type
Theory~\cite*{cohenCubicalTypeTheory2018,angiuliComputationalSemanticsCartesianCubical2019,vezzosiCubicalAgdaDependently2019}
as well.  In App.~\ref{app:identitytypes}, App.~\ref{app:homotopytypes}, and App~\ref{app:hits}, we review the basics of
identity types, homotopy types, and HITs and refer the reader to the book for more details.

@@

\begin{toappendix}
  \subsection{Identity Types}
  \label{app:identitytypes}

@@

  \noindent A homotopy between functions $f \htpy g$ is given by pointwise equality between them $\dfun{x:A}{f(x) \id_{B} g(x)}$.
  The identity type for functions is equivalent to homotopies between them ${(f \id_{A \to B} g)} \eqv {(f \htpy g)}$, by
  function extensionality. An equivalence between types $A \eqv B$ is given by a pair of functions between them which
  compose to the identity, $f \comp g \htpy \idfunc_{B}$ and $g \comp f \htpy \idfunc_{A}$, and this is equivalent to the
  identity type for the universe, $(A \id_{\UU} B) \eqv (A \eqv B)$, by univalence.

@@

  The type $\dsum*{x:A}{P(x)}$ is the collection of all the fibers and is called the total space of $P$. The first
  projection ${\pi_1 : \dsum*{x:A}{P(x)} \to A}$ from the total space to the base space $A$ has the structure of a
  fibration, that is, there is a lifting operation (\Cref{fig:lift}) which lifts paths in the base space to paths in the
  total space. Given a path $p : x \id_{A} y$ in the base space, and $u : P(x)$ a point in the fiber over $x$, we have:

@@

Functions between types are functors between groupoids, and type families (or functions to the universe) are indexed
families of groupoids. A type family $P : A \to \UU$ comes equipped with a functor $\pi_1 : \dsum*{x:A}{P(x)} \to \UU$
which has a lifting operation giving it the structure of a fibration. The
$\term{transport}$ operation $\transport{P}{p}{\blank}$ lifts paths in the base space to equivalences between the fibers:
% Using the groupoid structure of
% $A$, we can show that $\transport{P}{p}{\blank}$ and $\transport{P}{\inv{p}}{\blank}$ form an equivalence.
% We name
% this map $\tptEqv{P}$ which lifts paths to equivalences of fibers.
\[
  \tptEqv{P} : \dfun{x,y:A}{x \id_{A} y \to P(x) \eqv P(y)}
\]

\noindent The type families (or fibrations) we are interested in are the ones where paths in the base space completely determine
the equivalences in the fibers -- these are called univalent
fibrations~\cite*{kapulkinUnivalenceSimplicialSets2018,kapulkinSimplicialModelUnivalent2021,christensenCharacterizationUnivalentFibrations2015}.

@@

  \subsection{Homotopy Types}
  \label{app:homotopytypes}

@@

  \subsection{Higher Inductive Types}
  \label{app:hits}

@@

  \begin{definition}[Set Quotient]
    \label{def:set-quot}
    Given a type $A$ which is an $\hSet$, and a relation $R : A \to A \to \hProp$, the set-quotient $\quot{A}{R}$ is the
    higher inductive type generated by
    \begin{itemize}
      \item an inclusion function $q : A \to \quot{A}{R}$,
      \item for each $x, y : A$ such that $R(x,y)$, a path $q(x) \id_{\quot{A}{R}} q(y)$,
      \item a set truncation, for each $x, y : \quot{A}{R}$ and $r, s : x \id_{\quot{A}{R}} y$, we have $r \id s$,
    \end{itemize}
    with an appropriate induction principle.
  \end{definition}

@@

\begin{proposition}[Univalent Subuniverse]
  \label{prop:univsub}
  A universe predicate is a type family $P : \UU \to \UU$ whose fibers are propositions, that is, $P(X)$ is a
  proposition for every $X$. Given such a predicate, the fibration ${\fst : \dsum*{X:\UU}{P(X)} \to \UU}$ is univalent and generates a
  univalent subuniverse $\UU_{P} \defeq (\dsum{X:\UU}{P(X)}, \fst)$.
\end{proposition}

\begin{appendixproof}
  (of Prop.~\ref{prop:univsub})
  Suppose $(U, \El) \defeq (\dsum*{X:\UU}{P(X)}, \fst)$ is a subuniverse generated by a subtype $P : \UU \to \UU$. For
  any $X, Y : \UU$ such that $\phi : P(X)$ and $\psi : P(Y)$, we want to show that
  $\tptEqv{\fst} : (X,\phi) \id (Y,\psi) \to X \eqv Y$ is an equivalence. We construct
  $X \eqv Y \to (X,\phi) \id (Y,\psi)$ by $\ua$ and using the fact that $P(\blank)$ is a proposition. That it is an
  inverse follows by calculation using the appropriate computation rules.
\end{appendixproof}

The types we are interested in are the finite types. In constructive mathematics, the notion of finiteness is
subtle~\cite{spiwackConstructivelyFinite2010}. We use the notion of Bishop-finiteness: a type is finite if it is
merely equivalent to a finite set (\Cref{def:finite-set} and \Cref{def:isfin}).

\begin{definition}[$\Fin$]
  \label{def:finite-set}
  The type family $\Fin : \Nat \to \UU$ is the type of finite sets indexed by their cardinality. It is defined
  equivalently in two different ways,
  \begin{gather*}
    \begin{aligned}
      \Fin[n] & \defeq \dsum{k:\Nat}{k < n}
    \end{aligned}
    \qquad\qquad \qquad\qquad
    \begin{aligned}
       & \Fin[0] \defeq \bot                      \\
       & \Fin[\suc[n]] \defeq \top \sqcup \Fin[n]
    \end{aligned}
  \end{gather*}
  Note that $\Fin[n]$ is a set, and we use both definitions interchangeably.
\end{definition}
